---
title: "Lab 2 - Research Plan and Dataset Hopping"
output:
  html_notebook:
    toc: yes
    toc_depth: 3
    toc_float: yes
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
editor_options:
  chunk_output_type: inline
---

## Instructions and Overview

In this lab, you will begin to outline a research plan for a project analyzing data regarding social vulnerability in the wake of a pandemic. In your groups on Tuesday, you selected a topic for data investigation. This will be the topic you write about in this lab. At this point, we are just documenting what we currently know about our topic and what questions we have regarding the topic. You need not do external research in the first part of this assignment. In the second part of the assignment, you will begin searching for data related that could help address some of the empirical research questions that you devise. At the end of the course, we will come back to this lab to document how our understanding of the topic has changed following several weeks of working with data. 

## Research Plan

To begin your research plan, please respond to the questions below in full sentences. Again, at this point, you do not need to do external reserach on your topic. You are simply documenting what you currently know about your topic. 

### What topic am I interested in investigating?

```{r eval=FALSE}
Fill response here. 
```

### Why is this topic relevant to contemporary public health or social vulnerability in the wake of a disaster?

```{r eval=FALSE}
Fill response here. 
```

*Empirical research questions* are questions that an analyst can assess evidence to address. Examples of social-theoretical questions include:

* Do United States hospitals have enough beds to accommodate the expected influx in Covid-19 cases? At what point will there not be enough beds available?
* In which states are hospitals better equipped with beds and staff to take on an influx in patients?

### What empirical research questions do I intend to address in my research? List at least two.

```{r eval=FALSE}
Fill response here. 
```

*Social-Theoretical questions* are questions about how certain social, political, or environmental phenomena operate at a broad scale. We often cannot answer these questions with one research project. However, we often aim to increase understanding of these questions through a research project. Examples of social-theoretical questions include:

* How has the United States prioritized critical health infrastructure?
* How do national hospital business models implicate social vulnerability in the wake of a public health crisis?

### What social-theoretical questions do I intend to address in my research? List at least two. 

```{r eval=FALSE}
Fill response here. 
```

### What data do I believe exists about this topic?

Fill response in the table below:

Dataset | Do you think this data is accessible?
-------- | --------
Number of cases of Covid-19 globally  | Yes
Fill | Fill
Fill | Fill
Fill | Fill
Fill | Fill
Fill | Fill

### What are the key events, dates, or years relevant to this topic?

```{r eval=FALSE}
Fill response here. 
```

### What are the key sites, locations, or geographies relevant to this topic?

```{r eval=FALSE}
Fill response here. 
```

### What social groups are impacted by this topic?

```{r eval=FALSE}
Fill response here. 
```

### How is the environment implicated in this topic?

```{r eval=FALSE}
Fill response here. 
```

### How is the economy implicated in this topic?

```{r eval=FALSE}
Fill response here. 
```

### How are politics implicated in this topic?

```{r eval=FALSE}
Fill response here. 
```

### How am I implicated in this topic?

```{r eval=FALSE}
Fill response here. 
```

### What are some of the most common ways different social groups talk about this topic?

```{r eval=FALSE}
Fill response here. 
```

---

## Dataset Hopping

In May 2009, Data.gov - a web portal for accessing US government datasets - was launched by then federal Chief Information Officer Vivek Kundra. Following this, in December 2009, then US President Barack Obama signed the Open Government Data Directive, requiring that all federal agencies post at least 3 high value datasets on data.gov within 45 day. A few years later in May 2013, Pres. Obama signed an Executive Order to: "M"ak[e] Open and Machine Readable the New Default for Government Information." The Order required that the US Office of Management and Budgeting, in collaboration with the CIO and CTO, put out and oversee an Open Data Policy. This policy required the following:
* Data needs to be published in machine-readable formats
* Data needs to be described with Project Open Data metadata
* Data needs to be licensed openly
See [here](https://www.whitehouse.gov/sites/whitehouse.gov/files/omb/memoranda/2013/m-13-13.pdf ) for more details. 
Let's talk about what each of these requirments mean:

### Machine-readable

Machine-readable data is data that can be readily processed by a computer. Typically machine-readable data is structured in ways that many different computer applications can recognize. There are different degrees of machine-readability of digital data:

* Data stored in tables in **PDFs** will be the bain of our existence. This is perhaps some of the least machine-readable data that we will encounter in our data research. This is because data in tables in PDFs is not structured in such a way that computer software can easily reference and operate on specific values. 
* Data stored in **Excel** files is a bit more machine-readable. We can import data stored into an Excel file into Excel, and the software will recognize how the data is strucutured, aggregate values into separate cells, and thus present it to us in a way that we can operate on it. However, to open Excel, we need access to Microsoft products, which we have to pay for. 
* Data stored in **Comma Separated Value (of .csv)** files is more machine-readable than data stored in Excel files. A CSV file is a text file structured so that line in the file designates a records and various fields for describing that record are separated by commas. If we were to open a CSV file in a basic text editor, it would look something like this:

Name, Age, Birth Month, Time on Phone
Sally, 23, 3, 42
Julie, 40, 2, 98
Mark, 14, 8, 120

However, if we were to open the same file in Excel, each value would be seprated into its own cell. A CSV file is software independent. We can view it in just about any comptuer application that displays data. 

While we won't work with such formats in this course, data can be made even more machine-readable than a CSV file. Formats like XML and RDF allow us to strucutre data with much more specificity. They are often considered the gold standards of machine-readability.

### Metadata

Metadata is data about data, and Project Open Data Metadata is data about datasets. It is considered good practice to document datasets with metadata as it provides information about the context of the dataset to those that will view and use it. There are two kinds of metadata:
1. Administrative metadata: data about how a dataset is/should be managed
2. Descriptive metadata: data about what is inside a dataset

Administrative metadata will answer questions such as: 
* Who created this data?
* When did they create it?
* When was it last updated?
* How often will it be updated?
* Who published it?
* Who owns the data, and how is it licensed?

Why is administrative metadata so important? Here are just a few reasons:

* If we run into errors, we need to know who to contact.
* We need to know how often it is updated to get a sense of its timeliness. 
* We need to know how it is licensed so that we understand whether we have permission to use it, to distribute it, or to publish it.
* We need to know who created it in order to understand what perspective the data represents. 
* We need to be able to cite it. 

Descriptive metadata will answer questions such as:
* How many columns are in the dataset, and how many rows?
* What does each row represent?
* What does each column name refer to?

Why is descriptive metadata so important? Here are just a few reasons:

* We need to ensure that we don’t make improper assumptions of what each column means.
* We need to know the scope of the data.
  * What timeframe does it cover?
  * What geography does it cover?
  * What topics does it cover?
* We need to ensure that we know what each row represents. 
* We need to ensure that we know how each observation is defined. 
* We need to understand the range of values that might appear in a field.
* We need to know how missing fields get represented.
* We need to know about any known inconsistencies in the data. 

### Licensed Openly

Just because something is available on the Web does not necessarily mean that we are free to download and use it as we please. Historically, different government agecies would allow access to certain public datasets for a fee - a fundraising effort of the agency. (Oftentimes, we hear this referred to as data being behind a "paywall.")

At this point in the quarter, you will begin to search for open government or academic datasets. 

Questions

* Dataset?: What is the title of the dataset?
* Found?: Where did you find it?
* Formats?: In what formats is it available for download?
* DD?: Is there a data dictionary or data documentation available for download?
* Timespan?: What range of time does it cover?
* Geography? What geographic range does it cover? 
* Row?: What does each row in the dataset represent?
* Update?: How often is the dataset updated?

Dataset? | Found? | Formats? | DD? | Timespan? | Geography? | Row? | Update?
------- | ------- | ------- | ------- | ------- | ------- | ------- | -------
2019 Novel Coronavirus COVID-19 | https://github.com/CSSEGISandData/COVID-19 | .csv | No | Jan 22, 2020 to Present  | Global by Province/Country | Cases per day by Province/Country | Daily 
Fill | Fill | Fill | Fill | Fill | Fill | Fill | Fill
Fill | Fill | Fill | Fill | Fill | Fill | Fill | Fill
Fill | Fill | Fill | Fill | Fill | Fill | Fill | Fill
Fill | Fill | Fill | Fill | Fill | Fill | Fill | Fill


