---
title: "Lab 2 - Research Plan and Dataset Hopping"
output: 
  github_document:
    toc: true
    toc_depth: 2
editor_options: 
  chunk_output_type: inline
---

## Instructions and Overview

R Markdown is an authoring framework for R. It enables you to style and publish documents with code for data analysis embedded within it. In future labs, we will embed code in our documents. However for now, here are a few tricks for styling documents in R Markdown. 

### Code Chunks

```{r eval=FALSE}
This grey box between ```{r} and ``` is a code chunk. 

Typically, when we put code in this chunk and then publish our R Markdown document, 
the code written here will be run and output on the screen. 

However, by placing eval=FALSE in the brackets, we tell R not to evaluate the code. 

This is useful when we are just filling text into the code chunk.
```

### Headings

```{r eval=FALSE}
Use '#' to create headings:
  
# Heading 1
## Heading 2
### Heading 3
```

### Tables

```{r eval=FALSE}
Tables can be formated like this:
  
Header 1 | Header 1
-------- | --------
Cell 1  | Cell 2
Cell 3  | Cell 4
 
  
```

Header 1 | Header 1
-------- | --------
Cell 1  | Cell 2
Cell 3  | Cell 4

### Bullets and Numbering

```{r eval=FALSE}
You can bullet content using the '*':

* Point 1
* Point 2
```

This will output:

* Point 1
* Point 2

```{r eval=FALSE}
You can number content like this:

1. Point 1
2. Point 2
```

This will output:

1. Point 1
2. Point 2


For more information about formatting documents in R Markdown see [this](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf) cheatsheet. Note how I created a hyperlink here!

---

## Research Plan

### What topic am I interested in investigating?

```{r eval=FALSE}
Fill response here. 
```

### Why is this topic relevant to contemporary public health or social vulnerability in the wake of a disaster?

```{r eval=FALSE}
Fill response here. 
```

*Empirical research questions* are questions that an analyst can assess evidence to address. Examples of social-theoretical questions include:

* Do United States hospitals have enough beds to accommodate the expected influx in Covid-19 cases? At what point will there not be enough beds available?
* In which states are hospitals better equipped with beds and staff to take on an influx in patients?

### What empirical research questions do I intend to address in my research? List at least two.

```{r eval=FALSE}
Fill response here. 
```

*Social-Theoretical questions* are questions about how certain social, political, or environmental phenomena operate at a broad scale. We often cannot answer these questions with one research project. However, we often aim to increase understanding of these questions through a research project. Examples of social-theoretical questions include:

* How has the United States prioritized critical health infrastructure?
* How do national hospital business models implicate social vulnerability in the wake of a public health crisis?

### What social-theoretical questions do I intend to address in my research? List at least two. 

```{r eval=FALSE}
Fill response here. 
```

### What data do I believe exists about this topic?

Fill response in the table below:

Dataset | Do you think this data is accessible?
-------- | --------
Number of cases of Covid-19 globally  | Yes
Fill | Fill
Fill | Fill
Fill | Fill
Fill | Fill
Fill | Fill

### What are the key events, dates, or years relevant to this topic?

```{r eval=FALSE}
Fill response here. 
```

### What are the key sites, locations, or geographies relevant to this topic?

```{r eval=FALSE}
Fill response here. 
```

### What social groups are impacted by this topic?

```{r eval=FALSE}
Fill response here. 
```

### How is the environment implicated in this topic?

```{r eval=FALSE}
Fill response here. 
```

### How is the economy implicated in this topic?

```{r eval=FALSE}
Fill response here. 
```

### How are politics implicated in this topic?

```{r eval=FALSE}
Fill response here. 
```

### How am I implicated in this topic?

```{r eval=FALSE}
Fill response here. 
```

### What are some of the most common ways different social groups talk about this topic?

```{r eval=FALSE}
Fill response here. 
```

---

## Dataset Hopping

Questions

* Dataset?: What is the title of the dataset?
* Found?: Where did you find it?
* Formats?: In what formats is it available for download?
* DD?: Is there a data dictionary or data documentation available for download?
* Timespan?: What range of time does it cover?
* Geography? What geographic range does it cover? 
* Row?: What does each row in the dataset represent?
* Update?: How often is the dataset updated?

Dataset? | Found? | Formats? | DD? | Timespan? | Geography? | Row? | Update?
------- | ------- | ------- | ------- | ------- | ------- | ------- | -------
Novel Coronavirus (COVID-19) Cases Data | https://data.humdata.org/dataset/novel-coronavirus-2019-ncov-cases | .csv | No | Jan 22, 2020 to Present  | Global by State/Country | Cases per day by State/Country | Live 
Fill | Fill | Fill | Fill | Fill | Fill | Fill | Fill
Fill | Fill | Fill | Fill | Fill | Fill | Fill | Fill
Fill | Fill | Fill | Fill | Fill | Fill | Fill | Fill
Fill | Fill | Fill | Fill | Fill | Fill | Fill | Fill


