---
title: "Lab 3 - Context, Historical Background, and Stakeholders"
output: 
  github_document:
    toc: true
    toc_depth: 2
editor_options: 
  chunk_output_type: inline
---

## Instructions and Overview

This week you are going to do some background research on the dataset that you will be primarily working with throughout the quarter. As we are learning in our reading of *The Fires*, it is important that we understand how the datasets we analyze get produced, who produced them, what they are designed to represent (and also what they might ignore), and what socio-political contexts might affect what their scope. 

As you respond to these questions, you should be sure to examine the data documentation, the web source where you retreived the data, the website of the organization that collected or published the data, and any laws/press releases/news articles that reference the data. **All responses should be in your own words, and be sure to cite all sources.** You most likely will not be able to answer all questions below. This is OK. If cannot find an answer to a question, please describe the steps you took to try and find the answer. 

## Data Context and Historical Background

### Who collects the data? 

```{r eval=FALSE}
Fill response here. 
```

### Who aggregates the data?

```{r eval=FALSE}
Fill response here. 
```

### Who publishes the data?

```{r eval=FALSE}
Fill response here. 
```

### Why are these groups involved in the production of the data?

```{r eval=FALSE}
Fill response here. 
```

### How long has the data been collected? Have there been any major changes to how the data gets collected since its inception?

```{r eval=FALSE}
Fill response here. 
```

### According to what procedures is the data collected?

```{r eval=FALSE}
Fill response here. 
```

### Where is the data collected? How often is the data collected?

```{r eval=FALSE}
Fill response here. 
```

### Are there any laws that mandate the data to be collected? If so, what are they?

```{r eval=FALSE}
Fill response here. 
```

### Have any social movements pushed for the data to be collected? How have these groups exerted influence over the data collection?

```{r eval=FALSE}
Fill response here. 
```

### What technologies, algorithms, or other datasets are needed to produce this dataset?

```{r eval=FALSE}
Fill response here. 
```

### List any expenses that you can anticipate being involved in producing this data. Who pays for these expenses?

```{r eval=FALSE}
Fill response here. 
```

### What definitions, models, or indexes have shaped the way that numbers in this dataset get quantified?

```{r eval=FALSE}
Fill response here. 
```

### How long has this dataset been public? Under what circumstances was this dataset made public?

```{r eval=FALSE}
Fill response here. 
```

### Through what channels has the dataset been made available to the public?

```{r eval=FALSE}
Fill response here. 
```

### Do any institutions leverage the dataset to support decision-making? Which institutions, and which decisions?

```{r eval=FALSE}
Fill response here. 
```

### How has this dataset been talked about in the media?

```{r eval=FALSE}
Fill response here. 
```

### Which people, places, things, or issues may be over-represented in the data? Why?

```{r eval=FALSE}
Fill response here. 
```

### Which people, places, things, or issues may be under-represented in the data? Why?

```{r eval=FALSE}
Fill response here. 
```

---

## Stakeholders

Fill out the chart below for at least five of the datasets's stakeholders.

* Stakeholder?: Name a stakeholder for this dataset. Be specific. 
* Interest?: Characterize the interest that this stakeholder has in the dataset. 
* Motivation?: What motivates this interest in the data?
* Influence? Characterize the influence the stakeholder has over collection, aggregation, or publication of the data.


Stakeholder? | Interest? | Motivation? | Influence? 
------------ | ------------ | ------------ | ------------ 
Fill | Fill | Fill | Fill 
Fill | Fill | Fill | Fill 
Fill | Fill | Fill | Fill 
Fill | Fill | Fill | Fill 
Fill | Fill | Fill | Fill 


