---
title: "Lab 6 - Quantitative Insights"
output: 
  github_document:
    toc: true
    toc_depth: 2
editor_options: 
  chunk_output_type: inline
---

## Instructions and Overview

In this lab, we will practice calculating and interpreting measures of central tendency and measures of dispersion, as well as the consequences of relying on such numbers out of context to represent complex problems. To begin you will need to import and clean your dataset, and then you will follow the prompts while responding to short answer questions. Examples have been provided to support you throughout the process. At the end of the assignment, we will integrate some of your calculations into the Shiny application you started last week. 

## Getting Started

### Load the relevant libraries
```{r}
library(tidyverse)
library(lubridate)
library(shiny)
library(shinydashboard)
library(shinyWidgets)
```

### Import and clean example datasets 
```{r}
hospitals <- read.csv("https://raw.githubusercontent.com/lindsaypoirier/STS-115/master/datasets/Hospitals.csv", stringsAsFactors = FALSE)

cases <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv", stringsAsFactors = FALSE)

#Do not worry about this line of code for now. Since the cases data gets appended every day with a new column representing that day's case counts, if we want the total cases per country, we need to add up all of the previous day's counts into a new column. The column below does this for us. 
cases <- 
  cases %>% 
  mutate(Total.Cases = 
           cases %>% 
           select(starts_with("X")) %>% 
           rowSums()
         ) %>%
  select(Province.State, Country.Region, Total.Cases)

hospitals$ZIP <- as.character(hospitals$ZIP)

hospitals$ZIP <- str_pad(hospitals$ZIP, 5, pad = "0") 

is.na(hospitals) <- hospitals == "NOT AVAILABLE"
is.na(hospitals) <- hospitals == -999
is.na(cases) <- cases == ""

hospitals$SOURCEDATE <- ymd_hms(hospitals$SOURCEDATE)
hospitals$VAL_DATE <- ymd_hms(hospitals$VAL_DATE)
```

### Import and clean your dataset. 
```{r}
#Copy and paste relevant code from Lab 4 to import your data here. 

#Copy and paste relevant code from Lab 4 to clean your data here. This includes any row binding, character removals, converions in variable type, date formatting, or NA conversions. 
```

## Measures of Central Tendency

A measure of central tendency is a single numeric quantity describing data by identifying a central position within it. In this lab, we will calculate the mean and the median of values within our data. Under certain circumstances, one of these measurements can be more appropriate than the other for summarizing data. The frequency plots that we produced last week will help us to determine which measurement is more appropriate. Be sure to have them on hand to review as we move through this part of the assignment. 

### Mean

The most common measure of central tendency is a *mean* - the sum of a series of values divided by the number of values summed. This measure considers every value in a set of data and thus is a model of the entire set of data. Remember from lab four that we can do this with **summarize()** 

```{r}
#df %>% summarize(mean_value = mean(NUMERIC_VARIABLE, na.rm = TRUE))

hospitals %>% 
  summarize(mean_value = mean(BEDS, na.rm = TRUE))

#Fill your code here. 
```

> Note how we are choosing to ignore NA values above by calling na.rm = TRUE. When we do so, we need to keep in mind that we are not summarizing across all observations in the dataset, but only those for which there is a value listed in the variable we are operating on. Why might it be important to make communicate that we've made this move when we are presenting our data?

Now, let's calculate the mean of a numeric variable in your dataset. 

```{r}
#Fill your code here. 
```

### Median

Another measure of central tendency considers to the middle value in a dataset; this is referred to as the *median*. 

```{r}
#df %>% summarize(median_value = median(NUMERIC_VARIABLE, na.rm = TRUE))

hospitals %>% 
  summarize(median_value = median(BEDS, na.rm = TRUE))
```

Let's calculate the median of a numeric variable in your dataset. 

```{r}
#Fill your code here. 
```

You'll notice in the hospitals dataset, there's a considerable difference between the mean and the median value. So which should we rely on to summarize the central location in our dataset? 

We want to keep in mind that because the mean is a model of the entire dataset, it is easily influenced by outliers in the dataset, as well as a skewed distribution. When the distribution of values in a variable is normal, the mean and median of the dataset will be the same because the values on either side of the middle value will be symmetrical. In other words, the distribution will be balanced on either side of the median. Frequency plots give us a good indication of the distribution of values the dataset. 

```{r fig.height=5, fig.width=10}
hospitals %>% 
  ggplot(aes(x = BEDS)) +
  geom_freqpoly(binwidth = 10)
```
You'll notice in the hospitals dataset, the values of beds are significantly right-skewed. We can tell because the values represented on the plot are far from symmetrical. Instead, there is a long tail running to the right of the data, indicating that, while a considerable majority of hospitals have fewer than 500 beds available, there are at least a few hospitals that have up to 1600 beds available. In such cases, the mean is going to be significantly influenced by the larger values in the data even though there are far fewer. 

Create a frequency plot of the numeric variable you produced calculations of above. (You may have created this last week, and you may copy and paste it from the previous lab.)

```{r fig.height=5, fig.width=10}
#Create a frequency plot here. 
```

Characterize the distribution of values in the frequency plot you created. Is it symmetrical or skewed?

```{r eval=FALSE}
Fill your response here. 
```

When the values in a frequency plot are skewed to the left or the right, we often want to rely on the median rather than the mean as a measure of central tendency, as it more clearly distinguishes the central location in the data. In the hospitals dataset, median is a better indicator of central tendency than the mean. Which value is more appropriate as a measure of central tendency in your data? 

```{r eval=FALSE}
Fill your response here. 
```

Characterize the extent to which the value you selected as a measure of central tendency above (mean or median) is representative of your data. What story is told when we rely on this number as a summary of the data? If we were to use this single value to summarize this variable, what narratives would be left out?

```{r eval=FALSE}
Fill your response here. 
```

Note that we can see all of the values we calculated above with the summary() function we worked with in lab four. 

```{r}
#df %>% select(NUMERIC_VARIABLE) %>% summary()

hospitals %>% 
  select(POPULATION) %>% 
  summary()
```

Select another numeric variable in your dataset and call summary() on it.

```{r}
#Fill your code here. 
```

What story do these calculations tell? Would you select the median or the mean as a measure of central tendency? Why?

```{r eval=FALSE}
Fill your response here. 
```

Characterize the extent to which the value you selected as a measure of central tendency above is representative of your data. If we were to use this single value to summarize this variable, what narratives would be left out?

```{r eval=FALSE}
Fill your response here. 
```

As we've been discussing in class, any measure that we calculate is mediated by the way the phenomena we are measuring gets defined. How was the numeric variable that you selected defined? What is included in this definition, and what potentially relevant values are excluded?

```{r eval=FALSE}
Fill your response here. 
```

Who has stakes in how this variable gets defined? In what way might these stakes influence the measures that we have calculated?

```{r eval=FALSE}
Fill your response here. 
```

## Measures of Dispersion

Measures of dispersion help us to understand how spread out the values in our dataset are - their variations from each other and from the data's mean. 

### Ranges

We can also use summarize to find the max value, min value, and range of values in a numeric variable. 

```{r}
#df %>% summarize(max_value = max(NUMERIC_VARIABLE, na.rm = TRUE))
hospitals %>% 
  summarize(max_value = max(BEDS, na.rm = TRUE))

#df %>% summarize(min_value = min(NUMERIC_VARIABLE, na.rm = TRUE))
hospitals %>% 
  summarize(min_value = min(BEDS, na.rm = TRUE))

#df %>% summarize(range = max(NUMERIC_VARIABLE, na.rm = TRUE) - min(NUMERIC_VARIABLE, na.rm = TRUE))
hospitals %>% 
  summarize(range = max(BEDS, na.rm = TRUE) - min(BEDS, na.rm = TRUE))
```

Select a numeric variable in your dataset and calculate the maximum value, the minimum value, and the range of values within that variable.

```{r}
#Fill your code to calculate max here. 

#Fill your code to calculate min here. 

#Fill your code to calculate range here. 
```

Do the numbers surprise you? What insight can you draw from this calculation?

```{r eval=FALSE}
Fill your response here. 
```

### Quartile Deviation

Even though we are introducing the boxplot in relation to quartile deviation, it can be a tool for summarizing a number of quantitative insights in relation to a numerical variable in our dataset. Boxplots provide a visual representation of both measures of central tendency and measures of dispersion. The center line in a boxplot indicates the median of the dataset. The bottom of the box represents the 1st quartile - the value in the middle of the minimum and the median (or the the value at the 1st quarter position). The top of the box represents the 3rd quartile - the value in the middle of the median and the maximum (or the value at the 3rd quarter position). The whiskers include almost all of the data - indicating its range from minimum to maximum excluding outliers. The dots represent outliers. ggplot has a calculation for outliers that we need not go into in this course.

The further the 1st and 3rd quartile are from the median (or, in other words, the wider the box), the greater the *quartile deviation.* In general, a narrower box and whiskers indicates less dispersion in the data, while a wider box and whiskers indicates greater dispersion.  

```{r fig.height=5, fig.width=5}
#df %>% ggplot(aes(y = NUMERIC_VARIABLE)) + geom_boxplot() + theme_bw()

hospitals %>%
  ggplot(aes(y = BEDS)) +
  geom_boxplot() +
  theme_bw()
```

Let's create a boxplot for one of your numeric variables below.

```{r fig.height=5, fig.width=5}
#Fill your plot here. Be sure to add a title and labels to your plot. 
```

What insight can you draw from the plot you created?

```{r eval=FALSE}
Fill your response here. 
```

### Standard Deviation

*Standard deviation* calculates the extent of concentration of values around the mean. A higher standard deviation indicates that values are more dispersed from the mean, and a lower standard deviation indicates that values are more concentrated around the mean.  

```{r}
#df %>% summarize(sd_value = sd(NUMERIC_VALUE, na.rm = TRUE))

hospitals %>% 
  summarize(sd_value = sd(BEDS, na.rm = TRUE))

cases %>% 
  group_by(Country.Region) %>%
  summarize(country_total = sum(Total.Cases, na.rm = TRUE)) %>%
  ungroup() %>%
  summarize(sd_value = sd(country_total, na.rm = TRUE))
```

Let's calculate the standard deviation with your data. Note that you may need to group_by(), summarize a total and then ungroup() to ensure that you are performing this calculation across values reported at the same scale.

```{r}
#Fill your code here.
```

What insight can you draw from this calculation?

```{r eval=FALSE}
Fill your response here. 
```

Now that you have a sense of the dispersion of some of the values in your data, why do you believe the values are as concentrated or dispersed as they are? What do the measures of dispersion you calculated say about social, political, or economic life in the area the data is representing? You may need to do some research to answer this question. For instance, noting the dispersion of beds in the hospitals data, I may do a Web search for "Why do some hospitals in the US have more beds than others?" and find [this](https://www.nytimes.com/interactive/2020/03/17/upshot/hospital-bed-shortages-coronavirus.html) article. Be sure to assess the reputability of your source and cite it in your response below. 

```{r eval=FALSE}
Fill your response here. 
```

## Grouped Quantitative Insights 

Like we learned in regards to co-variation, some of the most interesting information we can gain from our data is how values change depending on where in the data we are looking. 

ggplot's geom_boxplot feature is particularly good at comparing measures of dispersion across grouped values in a categorical variable. It is set up to visualize several boxplots side-by-side. Let's take a look at how we could compare dispersions in the number of hospital beds available by hospital ownership. 

```{r fig.height=5, fig.width=10}
#df %>% ggplot(aes(x = CATEGORICAL_VARIABLE, y = NUMERIC_VARIABLE)) + geom_boxplot()

hospitals %>%
  ggplot(aes(x = OWNER, y = BEDS)) +
  geom_boxplot() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + #Changes x-axis tick labels 90 degrees
  coord_flip() #Flips the x and y axis to make the data easier to read and compare
```

We see above that there is a greater dispersion in the number of beds available at state hospitals, non-profit hospitals, and federal hospitals. We also see that, in local hospitals, the median is very close to the first quartile, indicating that the data is going to be considerably right-skewed. Compare this to federal hospitals, where the median is more centered in the box (and there are few outliers). We are likely to see a bit more of a normal distribution in this data. Let's confirm this.

```{r fig.height=5, fig.width=5}
hospitals %>%
  filter(OWNER == "GOVERNMENT - LOCAL") %>%
  ggplot(aes(x = BEDS)) +
  geom_freqpoly(binwidth = 10) +
  theme_bw()

hospitals %>%
  filter(OWNER == "GOVERNMENT - FEDERAL") %>%
  ggplot(aes(x = BEDS)) +
  geom_freqpoly(binwidth = 10) +
  theme_bw()
```

Create a grouped boxplot for your dataset below.

```{r fig.height=5, fig.width=10}
#Fill your code here. Add a title and labels to your plot. 
```

Summarize what you learn from the plot. 

```{r eval=FALSE}
Fill your response here. 
```

How does the story that this grouped boxplot tells differ from the story told by the single boxplot you created above? Interpret why the story differs.

```{r eval=FALSE}
Fill your response here. 
```

Sometimes, we also want to graphically represent a calculation that we perform for a grouped categorical variable. For instance, let's say I wanted to show the median number of beds available grouped by the hospital owner. To do this, I need to calculate the median of the BEDS variable for each OWNER. One way to do this is to group_by(OWNER), summarize the median of BEDS, and then plot the resulting data frame. We discussed how to group_by() and summarize() in lab 4, and a ggplot call can be appended to these functions with a pipe (%>%) to plot them. However, here we can also use stat_summary(), in conjunction with a function. We will plot the median as a column plot by setting the geom = "col".

```{r fig.height=5, fig.width=10}
#df %>% ggplot(aes(x = CATEGORICAL_VARIABLE, y = NUMERIC_VARIABLE)) + stat_summary(geom = "col", fun.y = "median")

hospitals %>%
  ggplot(aes(x = OWNER, y = BEDS)) +
  stat_summary(geom = "col", fun.y = "median") +
  theme_bw() + # To change the plot theme
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + # To turn x-axis ticks 90 degrees
  scale_y_continuous(labels = scales::comma)  # To set x-axis ticks to comma notation
```

We can apply functions other than median in stat_summary(). Let's graphically represent the total number of beds for each owner. 

```{r fig.height=5, fig.width=10}
hospitals %>%
  ggplot(aes(x = OWNER, y = BEDS)) +
  stat_summary(geom = "col", fun.y = "sum") + #fun.y tells ggplot to perform this function on the y-variable
  theme_bw() + # To change the plot theme
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + # To turn x-axis ticks 90 degrees
  scale_y_continuous(labels = scales::comma)  # To set x-axis ticks to comma notation
```

I can group these calculations further using the *fill* aesthetic. Below we create three plots by setting the fill aesthetic to the hospitals' TYPE. Depending on the *position* attribute, we can create either a stacked column plot, a dodged column plot, or a filled column plot (which shows proportions of totals versus actual totals. )

```{r fig.height=5, fig.width=10}
#Stacked column plot of the total hospital beds by owner and type
hospitals %>%
  ggplot(aes(x = OWNER, y = BEDS, fill = TYPE)) +
  stat_summary(geom = "col", fun.y = "sum") +
  theme_bw() + # To change the plot theme
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + # To turn x-axis ticks 90 degrees
  scale_y_continuous(labels = scales::comma)  # To set x-axis ticks to comma notation

#Dodged column plot of the total hospital beds by owner and type
hospitals %>%
  ggplot(aes(x = OWNER, y = BEDS, fill = TYPE)) +
  stat_summary(geom = "col", fun.y = "sum", position = "dodge") +
  theme_bw() + # To change the plot theme
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + # To turn x-axis ticks 90 degrees
  scale_y_continuous(labels = scales::comma)  # To set x-axis ticks to comma notation

#Filled column plot of the total hospital beds by owner and type
hospitals %>%
  ggplot(aes(x = OWNER, y = BEDS, fill = TYPE)) +
  stat_summary(geom = "col", fun.y = "sum", position = "fill") +
  theme_bw() + # To change the plot theme
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + # To turn x-axis ticks 90 degrees
  scale_y_continuous(labels = scales::comma)  # To set x-axis ticks to comma notation

```

Using stat_summary() in conjunction with a function, create three column plots below. If using the sum function, experiment with the fill and the position. 

```{r fig.height=5, fig.width=10}
#Create your first plot here. Be sure to add a title and labels. 
```

```{r fig.height=5, fig.width=10}
#Create your second plot here. Be sure to add a title and labels. 
```

```{r fig.height=5, fig.width=10}
#Create your third plot here. Be sure to add a title and labels. 
```

Summarize what you learn from these three plots. 

```{r eval=FALSE}
Fill your response here. 
```

Which social groups might be interested in each of the plots that you produced? Why would they be interested in these plots?

```{r eval=FALSE}
Fill your response here. 
```

## Continue your shiny app.

Now we will incorporate some of the values we calculated and some of the plots we produced above into the Shiny App. In the server function, follow the instructions in the comments to create three value boxes in your app. You will also place two more plots in the app. Don't worry about styling the app. That will be the focus of a later lab. 

```{r}
ui <- dashboardPage(
  
  dashboardHeader(title = "TITLE HERE"),
  
  dashboardSidebar(
    #inputs will go here. 
  ),
  
  dashboardBody(
      infoBoxOutput("value1", width = 4),
      infoBoxOutput("value2", width = 4),
      infoBoxOutput("value3", width = 4),
      box(plotOutput("plot1")),
      box(plotOutput("plot2")),
      box(plotOutput("plot3")),
      box(plotOutput("plot4"))
  )
)
  

```


```{r}
server <- function(input, output, session) {
  
  output$value1 <- renderInfoBox({
    quant_insight1 <- 0
    #Replace '0' above with the code for one of the values you calculated above. Replace 'FILL DESCRIPTION HERE' with a brief description of this number.  
    infoBox(quant_insight1,'FILL DESCRIPTION HERE', icon = icon("stats", lib='glyphicon'), color = "purple")
  })
  
  output$value2 <- renderInfoBox({
    quant_insight2 <- 0
    #Replace '0' above with the code for one of the values you calculated above. Replace 'FILL DESCRIPTION HERE' with a brief description of this number.  
    infoBox(quant_insight2,'FILL DESCRIPTION HERE', icon = icon("stats", lib='glyphicon'), color = "purple")
  })
  
  output$value3 <- renderInfoBox({
    quant_insight3 <- 0
    #Replace '0' above with the code for one of the values you calculated above. Replace 'FILL DESCRIPTION HERE' with a brief description of this number.  
    infoBox(quant_insight3,'FILL DESCRIPTION HERE', icon = icon("stats", lib='glyphicon'), color = "purple")
  })
  
  
  output$plot1 <- renderPlot({
    hospitals %>% ggplot(aes(x = TYPE)) + geom_bar()
    #Replace plot above with your own plot. 
    
  })
  
  output$plot2 <- renderPlot({
    hospitals %>% ggplot(aes(x = TYPE)) + geom_bar()
    #Replace plot above with your own plot. 
  })
  
  output$plot3 <- renderPlot({
    hospitals %>% ggplot(aes(x = TYPE)) + geom_bar()
    #Replace plot above with your own plot. 
    
  })
  
  output$plot4 <- renderPlot({
    hospitals %>% ggplot(aes(x = TYPE)) + geom_bar()
    #Replace plot above with your own plot. 
  })
  
}
```


```{r}
shinyApp(ui, server)
```

